---
title: "CSCI E-63C Week 10 assignment"
output: html_document
---

```{r setup, include=FALSE}
library(randomForest)
library(MASS)
library(class)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this assignment we will compare performance of random forest to that of LDA and KNN on a simulated dataset where we know exactly what is association between predictors and outcome.  The relationship between predictor levels and the outcome will involve interaction that is notoriously difficult to model by methods such as LDA. The following example below illustrates the main ideas on a 3D dataset with two of the three attributes associated with the outcome:

```{r}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 1
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)
table(classTmp)
# plot resulting attribute levels colored by outcome:
pairs(xyzTmp,col=as.numeric(classTmp))
```

We can see that it is the interaction between the first two variables that has influences the outcome (we simulated it this way, of course!) and that points belonging to each of the two classes cannot be readily separated by a single line in 2D (or a single surface in 3D).

```{r}
# Split data into train and test
bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl
```

Random forest seems to do reasonably well on such dataset.

```{r}
# Fit LDA model to train data and evaluate error on the test data:
ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl
```

LDA, on the other hand, not so good! (not a surprise given what we've seen above).  What about a more flexible method such a KNN?  Let's check it out remembering that k -- number of neihbors -- in KNN is the parameter to modulate its flexibility (i.e. bias-variance tradeoff).

```{r}
# Fit KNN model at several levels of k:
dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate")
```

We can see from the above that there is a range of $k$ values where test error of KNN is the lowest and it is even lower that that of RF.  Now would be a good moment to think why one would want to choose RF over KNN or vice a versa for modeling the data if the figure above was representative of their true relative performance on a new dataset.

For the purposes of the assignment you can use the code above (probably best to wrap reusable parts of it into function(s)) to generate data with varying numbers of predictors associated with outcome and not, different numbers of observations and differences in the average values of predictors' between two classes as required below. These differences between datasets and parameters of the call to random forest will illustrate some of the factors influencing relative performance of random forest, LDA and KNN classifiers.  When comparing to KNN performance, please choose value(s) of `k` such that it performs sufficiently well -- feel free to refer to the plot above to select useful value(s) of `k` that you would like to evaluate here.  Keep in mind also that the value of `k` cannot be larger than the number of observations in the training dataset.

# Sub-problem 1 (15 points): effect of sample size

Generate datasets with `nObs=50`, `200` and `1000` observations (approximately evenly split between training and test datasets), two variables associated with the outcome as parameterized above and three not associated, and average difference between two classes same as above (i.e. in the notation from the above code `nClassVars=2`, `nNoisevars=3` and `deltaClass=1`).  Obtain random forest, LDA and KNN test error rates.  Describe the differences between different methods and across the sample sizes used here.

```{r p1_1000, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=1000")

```

```{r p1_200, echo=T}
# How many observations:
nObs <- 200
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:25)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=200")

```

```{r p1_50, echo=T}
# How many observations:
nObs <- 50
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:21)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=50")

```

> RF is always better than LDA. 

> RF and LDAperforms better with more data, RF is more obvious. 

> At about k=10 is the optimal point for KNN for all sample sizes where it can beat RF. 

# Sub-problem 2 (15 points): effect of signal magnitude

For sample sizes of `nObs=200` and `1000` observations (approximately evenly split into training and test datasets) simulate data as shown above with average differences between the two classes that are same as above, half of that and twice that (`deltaClass=0.5`, `1` and `2`).  Obtain and plot test error rates of random forest, LDA and KNN for each of the six (two samples sizes times three signal magnitudes) combinations of sample size and signal strengths.  Describe the most pronounced differences across error rates for those datasets: does the increase in the number of observations impact the error rate of the models?  Does change in the magnitude of signal impact their performance?  Are different classifier approaches impacted in a similar way?

```{r p2_200_05, echo=T}
# How many observations:
nObs <- 200
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 0.5
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:25)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=200 deltaClass=0.5")

```

```{r p2_200_1, echo=T}
# How many observations:
nObs <- 200
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:25)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=200 deltaClass=1")

```

```{r p2_200_2, echo=T}
# How many observations:
nObs <- 200
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 2
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:25)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=200 deltaClass=2")

```

```{r p2_1000_05, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 0.5
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=1000 deltaClass=0.5")

```

```{r p2_1000_1, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=1000 deltaClass=1")

```

```{r p2_1000_2, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 2
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nobs=1000 deltaClass=2")

```

> Larger observation size lowers error rate. 

> Increasing magnitude of signal lowers error rate. 

> All classifiers are impacted in the same way. 


# Sub-problem 3 (15 points): varying counts of predictors

For all possible pairwise combinations of the numbers of variables associated with outcome (`nClassVars=2` and `5`) and those not associated with the outcome (`nNoiseVars=1`, `3` and `10`) -- six pairwise combinations in total -- obtain and present graphically test errors from random forest, LDA and KNN.  Choose signal magnitude (`deltaClass`) so that it will yield non-trvial results -- noticeable variability in the error rates across those six pairwise combinations of attribute counts.  Describe the results: what is the impact of the increase of the number of attributes associated with the outcome on the classifier performance?  What about the number of attributes not associated with outcome - does it affect classifier error rate?  Are different classifier methods affected by these simulation parameters in a similar way?

```{r p3_2_1, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 1
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nClassVars=2, nNoiseVars=1")

```

```{r p3_2_3, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nClassVars=2, nNoiseVars=3")

```

```{r p3_2_10, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 2
# How many predictors are not:
nNoiseVars <- 10
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nClassVars=2, nNoiseVars=10")

```

```{r p3_5_1, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 5
# How many predictors are not:
nNoiseVars <- 1
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nClassVars=5, nNoiseVars=1")

```

```{r p3_5_3, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 5
# How many predictors are not:
nNoiseVars <- 3
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nClassVars=5, nNoiseVars=3")

```

```{r p3_5_10, echo=T}
# How many observations:
nObs <- 1000
# How many predictors are associated with outcome:
nClassVars <- 5
# How many predictors are not:
nNoiseVars <- 10
# To modulate average difference between two classes' predictor values:
deltaClass <- 1
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain])
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}
ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate nClassVars=5, nNoiseVars=10")

```

> Increasing number of attributes associated with outcome results in worse performance. 

> Increasing number of attributes not associated with outcome results in slightly worse performance, but ignorable. 

> The increase in number of attributes associated with outcome impacts all 3 classifiers similarly. And it impacts KNN more severely by shifting its optimal k. 

# Sub-problem 4: (15 points): effect of `mtry`

Parameter `mtry` in the call to `randomForest` defines the number of predictors randomly chosen to be evaluated for their association with the outcome at each split (please see help page for `randomForest` for more details).  By default for classification problem it is set as square root of the number of predictors in the dataset.  Here we will evaluate the impact of using different values of `mtry` on the error rate by random forest.

For `nObs=5000`, `deltaClass=2`, `nClassVars=3` and `nNoiseVars=20` generate data using the above approach and run `randomForest` on it with `mtry=2`, `5` and `10`.  Describe the impact of using different values of `mtry` on the error rate by random forest and compare it to that by LDA/KNN. 

```{r p4, echo=T}
# How many observations:
nObs <- 5000
# How many predictors are associated with outcome:
nClassVars <- 3
# How many predictors are not:
nNoiseVars <- 20
# To modulate average difference between two classes' predictor values:
deltaClass <- 2
# Simulate dataset with interaction between attribute levels associated with the outcome:
xyzTmp <- matrix(rnorm(nObs*(nClassVars+nNoiseVars)),nrow=nObs,ncol=nClassVars+nNoiseVars)
classTmp <- 1
for ( iTmp in 1:nClassVars ) {
  deltaTmp <- sample(deltaClass*c(-1,1),nObs,replace=TRUE)
  xyzTmp[,iTmp] <- xyzTmp[,iTmp] + deltaTmp
  classTmp <- classTmp * deltaTmp
}
classTmp <- factor(classTmp > 0)

bTrain <- sample(c(FALSE,TRUE),nrow(xyzTmp),replace=TRUE)

ldaRes <- lda(xyzTmp[bTrain,],classTmp[bTrain])
ldaTmpTbl <- table(classTmp[!bTrain],predict(ldaRes,newdata=xyzTmp[!bTrain,])$class)
ldaTmpTbl

dfTmp <- NULL
for ( kTmp in floor(1.2^(1:33)) ) {
  knnRes <- knn(xyzTmp[bTrain,],xyzTmp[!bTrain,],classTmp[bTrain],k=kTmp)
  tmpTbl <- table(classTmp[!bTrain],knnRes)
  dfTmp <- rbind(dfTmp,data.frame(err=1-sum(diag(tmpTbl))/sum(tmpTbl),k=kTmp))
}

```


```{r p4_2, echo=T}

# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain],mtry = 2)
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl
1-sum(diag(rfTmpTbl))/sum(rfTmpTbl)

ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate mtry=2")

```

```{r p4_5, echo=T}
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain],mtry = 5)
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl
1-sum(diag(rfTmpTbl))/sum(rfTmpTbl)

ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate mtry=5")

```

```{r p4_10, echo=T}
# Fit random forest to train data, obtain test error:
rfRes <- randomForest(xyzTmp[bTrain,],classTmp[bTrain],mtry = 10)
rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
rfTmpTbl
1-sum(diag(rfTmpTbl))/sum(rfTmpTbl)

ggplot(dfTmp,aes(x=k,y=err))+geom_point()+scale_x_log10()+geom_hline(aes(yintercept = err,colour=type),data=data.frame(type=c("LDA","RF"),err=c(1-sum(diag(ldaTmpTbl))/sum(ldaTmpTbl),1-sum(diag(rfTmpTbl))/sum(rfTmpTbl))))+ggtitle("KNN error rate mtry=10")

```

> Increasing mtry lowers error rate, but not very significant. 

> KNN performs far better than LDA and RF when there are a lot more noises than signals. 